Expression QTL hotspot dissection

### 2019-06-18 analysis/paper/paper.Rnw
added lists of figs & tables; removed references to nonexistent figures

\doublespacing
\tableofcontents
%\listoffigures
%\listoftables
%\pagestyle{headings} % page numbers appear at bottom of page, centered!
%\markright{\hfill}
\chapter{Introduction}

@@

Due to the breeding scheme, each Diversity Outbred mouse is a highly
heterozygous and essentially unique mosaic of
founders' DNA. Figure~\ref{fig:do} presents color-coded chromosomes for a collection of
subjects from a multiparental population like the Diversity Outbred mice. DNA from each founder
line is color-coded. Each pair of vertical bars corresponds to one subject. Each row corresponds to a single generation, with generation labels on the left. Generation number increases from top to bottom.
With each new generation, the Diversity Outbred mouse genomes are shuffled
in smaller and smaller chunks.

@@

\citet{crick1958protein} articulated a pathway for transmission of biological information
that is now known as the central dogma of molecular biology (Figure~\ref{fig:dogma})
\citep{crick1970central}.
In it, he argued that information encoded in DNA sequence is transmitted via transcription
to
RNA molecules, which, in turn, transfer the information to proteins via translation.
The process of transcription uses DNA as a template for creating a RNA molecule that
conveys the information encoded in DNA.
In this sense, every gene leads to a unique RNA molecule.
Translation is the molecular biology process by which an RNA molecule's information is
transferred to a protein.
As in transcription, the nucleic acid (DNA in transcription and RNA in translation) serves
as a template for synthesis of the product (RNA in transcription and protein in translation).
Thus, RNA molecules from distinct genes lead to different proteins.

@@

We continue by stating what it means for one trait to mediate a relationship between a
DNA variant and another trait. To clarify our discussion, we refer to an example from
\citet{chick2016defining} (Figure~\ref{fig:Dhtkd1}). \citet{chick2016defining}, in
studying livers of 192 Diversity Outbred mice, found evidence that \emph{Dhtkd1}
transcript levels associated with a Chromosome 2 marker near the \emph{Dhtkd1} gene.
They also found that
the same marker affected DHTKD1 protein concentrations. (Note that DHTKD1 protein is the
product of translation of \emph{Dhtkd1} transcripts.)
As anticipated, mediation analysis, in which the DHTKD1 protein concentrations are
regressed on founder allele dosages (at the Chromosome 2 marker) demonstrated that
\emph{Dhtkd1} transcript levels act as a mediator between DHTKD1 protein concentrations
and founder allele dosages. In fact, the extent of the reduction in association
strength indicates that the primary pathway by which the genetic marker affects DHTKD1
protein concentrations is through \emph{Dhtkd1} transcript levels.

@@

To determine whether a local transcript level mediates the relationship between
a nonlocal transcript level and a DNA variant, we perform a series of regression analyses,
which we detail below (Frame~\ref{frame3}).
In brief, we regress the nonlocal transcript levels on founder allele dosages at the DNA
variant, with and without conditioning on the candidate mediator (the local transcript levels).
If the LOD score measuring association between nonlocal trait and QTL genotype diminishes
sufficiently upon conditioning on a candidate,
then we declare the candidate a mediator.

@@

\begin{frameenv}{Four regressions for a single mediation analysis}\label{frame3}

@@

%\begin{equ}[!ht]
\begin{equation}
Y = XB + WC + M\beta + E
\label{model4}
\end{equation}
%\caption{Linear model with founder allele dosages, covariates, and candidate mediator.}
%\end{equ}
\end{frameenv}

We now consider the procedures needed for a mediation analysis in systems genetics.
In the four linear regression models (Frame~\ref{frame3}), $X$ is an $n$ by $8$ matrix of
founder allele dosages at a single marker, $B$ is an $8$ by $1$ matrix of founder allele
effects, $E$ is an $n$ by $1$ matrix of random errors, $b$
is a number, $1$ is
an $n$ by $1$ matrix with all entries set to $1$, $Y$ is
an $n$ by $1$ matrix of
phenotype values (for a single trait), and $M$ is an $n$
by $1$ matrix of values for a putative mediator.
$C$ is a $c$ by 1 matrix of covariate effects, and $W$ is
a $n$ by $c$ matrix of covariates.
We denote the coefficient of the mediator by $\beta$.

@@

\begin{frameenv}{Four assumptions for causal inference}\label{frame1}
  \begin{enumerate}
\item No unmeasured confounding of the DNA variant-nonlocal transcript levels relationship
\item No unmeasured confounding of the local transcript levels-nonlocal transcript levels relationship
\item No unmeasured confounding of the DNA variant-local transcript levels relationship
\item No local transcript levels-nonlocal transcript levels confounder that is affected by the DNA variant
\end{enumerate}

\end{frameenv}

Our first statistical challenge is due to confounding. To claim that the LOD
difference statistic reflects a causal relationship, four assumptions about
confounding are needed (Frame~\ref{frame1}) \citep{vanderweele2015explanation},
yet
it is often difficult or impossible to recognize unmeasured confounders. Unmeasured
confounders, as I use the term here, are unmeasured random variables that confound a causal
relationship between two other random variables.
For example, Figure~\ref{fig:confounding} shows a directed graph. Arrows indicate causal
directions. For example, ``QTL'' causes ``M''. ``M'' designates the mediator (local transcript). ``Y'' denotes the nonlocal transcript
level in this setting, while ``C1'', ``C2'', and ``C3'' are covariates that confound,
respectively, the DNA variant - nonlocal transcript level relationship, the local transcript
level - nonlocal transcript level relationship, and the DNA variant - local transcript level
relationship. Failure to measure C1, C2, or C3 leads to violations of Assumption 1,
Assumption 2, or Assumption 3, respectively (Frame~\ref{frame1}).

In studies of Diversity Outbred mice, relatedness is a possible confounder,
yet the linear models (Equations~\ref{model1},~\ref{model2},~\ref{model3},~\ref{model4}
in Frame~\ref{frame3}) fail to account for the complex relatedness patterns
among Diversity Outbred mice.
In fact, the only covariates in our models are for wave membership and sex.
The 400 mice were shipped from the Jackson Laboratory in four waves of 100 mice.
Because each wave was processed at a different time, I include wave membership
as a covariate with the goal of accounting for wave-related effects.
Other sources of confounding, such as batch effects in the phenotyping, may be
unmeasured.

@@

We designed a study to examine the possible roles for mediation analysis and pleiotropy testing.
Because \citet{keller2018genetic} reported that some nonlocal traits that map to
the Chromosome 2 hotspot did not demonstrate evidence of mediation by \emph{Hnf4a} expression
levels, we elected to study a collection of local gene expression traits,
rather than \emph{Hnf4a} alone.
This strategy enabled us to ask whether one of twelve other local traits mediates those
nonlocal hotspot traits that are not mediated by \emph{Hnf4a}.
Our set of local gene expression traits includes \emph{Hnf4a} and 12
other local genes (Table \ref{tab:annot}).
Our 13 local genes are the only genes that met three criteria (Frame~\ref{frame2}).
The 147 nonlocal traits that we studied all had LOD peak heights above 7.18 and QTL positions within 2 Mb of the center of the hotspot (at 165.5 Mb).

\begin{frameenv}{Local gene inclusion criteria}\label{frame2}
\begin{enumerate}
    \item QTL peak with LOD $>$ 40
    \item QTL peak position within 2 Mb of hotspot center (165.5 Mb)
    \item Gene midpoint within 2 Mb of hotspot center (165.5 Mb)
\end{enumerate}
\end{frameenv}

@@

Our bivariate QTL analyses involved the same 13 local expression traits and 147 nonlocal expression traits.
We described above (Frame~\ref{frame2}) the criteria for choosing these expression traits.

@@

We also examined the 1,911 pairs by considering one nonlocal gene at a time.
For each of the 147 nonlocal genes, I plotted LOD difference proportion
against pleiotropy test statistic values.
I present below (Figure \ref{fig:4nonlocal}) examples that illustrate some of the
observed patterns between pleiotropy test statistics and LOD difference proportion values.

@@

%\subsection{Scatter plot for all 1911 pairs}
Figure \ref{fig:lod-diff-prop-v-lrt-all} is a scatterplot of the results from all pleiotropy tests and mediation analyses.
Each point contains one local expression trait and one nonlocal expression trait.
I see that points with high values of LOD difference proportion tend to have small
values of pleiotropy test statistic, and those points with high values of the
pleiotropy test statistic tend to have small values of LOD difference proportion.

Some points demonstrate large values of pleiotropy test statistic (\emph{eg.}, $>$ 10),
yet still have sizeable LOD difference proportion statistics (Figure~\ref{fig:lod-diff-prop-v-lrt-all}).
One possible explanation for such points is that the univariate LOD (LOD$_1$) value is small,
so that even a modest LOD difference gives rise to a sizeable LOD difference proportion value.
A second possible explanation is that the phenotype variances for both phenotypes in the pair
are large enough to skew the null distribution of the test statistic towards larger values.
To distinguish between these two, I would
examine the LOD difference statistics and obtain bootstrap $p$-values.

In Figure~\ref{fig:lod-diff-prop-v-lrt-all}, I colored red the points that involve \emph{Hnf4a};
points with other local genes are blue.
The most striking feature of the coloring is that many red points have small
values of the pleiotropy test statistics and very high values of LOD difference
proportion. This is what I expect when \emph{Hnf4a} mediates the relationship
between a nonlocal trait and a QTL; namely, \emph{Hnf4a} and a nonlocal trait
share a pleiotropic QTL and conditioning on \emph{Hnf4a} diminishes the LOD
score for the association between nonlocal trait and QTL.

@@

To more thoroughly examine the pleiotropy test and mediation analysis
relationships across the 13 local genes,
I created 13 plots of LOD difference proportion against pleiotropy test
statistic (Figures~\ref{fig:hnf4a} and~\ref{fig:nothnf4a-12}).
They reveal common patterns.
First, I see no points in the upper right quadrant of each plot (Figure~\ref{fig:nothnf4a-12}).
This tells us that those nonlocal genes with high values of pleiotropy test statistic
(when paired with the specified local gene) have low values of LOD difference proportion.
Similarly, those nonlocal genes with high values of LOD difference proportion tend
to have small values of the pleiotropy test statistic.
Finally, some trait pairs demonstrate low values of both
the LOD difference proportion and pleiotropy test statistic.
This observation suggests that, for a given local expression trait, the nonlocal trait
is not mediated by the local expression trait yet it shares a pleiotropic locus.
Low power to resolve multiple nearby QTL may give rise to such findings.

In comparing the \emph{Hnf4a} plot (Figure \ref{fig:hnf4a}) with the other
12 plots (Figure \ref{fig:nothnf4a-12}),
I see that none of the 12 plots in Figure \ref{fig:nothnf4a-12} closely resembles Figure \ref{fig:hnf4a}.
\emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22} are closest in appearance to the plot of \emph{Hnf4a}.
However, each of \emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22}
has very few points with LOD difference proportion above 0.5,
while \emph{Hnf4a} has many points with LOD difference proportion above 0.5.

@@

Now that I've examined our results from the perspective of every one of 13 local gene expression traits,
I turn attention to the nonlocal gene expression trait perspective.
I present four nonlocal gene expression trait plots in Figure~\ref{fig:4nonlocal}. I chose
these four from the 147 to demonstrate two of the observed patterns. In a single scatterplot,
I've plotted 13 points, one for each local gene expression trait, paired with a single
nonlocal gene expression trait. Local expression trait \emph{Hnf4a} is colored in red,
while the other local genes are colored in blue. The top row of Figure~\ref{fig:4nonlocal}
features scatterplots for nonlocal expression traits \emph{Abcb4} and \emph{Smim5}. I see
that local expression trait \emph{Hnf4a} has, by far, the greatest value of LOD difference
proportion. In the second row of Figure~\ref{fig:4nonlocal}, I see a different pattern in
the plots for nonlocal expression traits \emph{Tmprss4} and \emph{Gm3095}. Both demonstrate,
for \emph{Hnf4a}, small values for the pleiotropy test statistic and small values of LOD
difference proportion. This tells me that \emph{Hnf4a} shares a pleiotropic locus with both
\emph{Tmprss4} and \emph{Gm3095}, yet none of the 13 local gene expression traits mediates
the relationship between these two nonlocal traits and the QTL.

@@

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

>>>>

\doublespacing
\tableofcontents
\listoffigures
\listoftables
%\pagestyle{headings} % page numbers appear at bottom of page, centered!
%\markright{\hfill}
\chapter{Introduction}

@@

Due to the breeding scheme, each Diversity Outbred mouse is a highly
heterozygous and essentially unique mosaic of
founders' DNA. Figure presents color-coded chromosomes for a collection of
subjects from a multiparental population like the Diversity Outbred mice. DNA from each founder
line is color-coded. Each pair of vertical bars corresponds to one subject. Each row corresponds to a single generation, with generation labels on the left. Generation number increases from top to bottom.
With each new generation, the Diversity Outbred mouse genomes are shuffled
in smaller and smaller chunks.

@@

\citet{crick1958protein} articulated a pathway for transmission of biological information
that is now known as the central dogma of molecular biology (Figure)
\citep{crick1970central}.
In it, he argued that information encoded in DNA sequence is transmitted via transcription
to
RNA molecules, which, in turn, transfer the information to proteins via translation.
The process of transcription uses DNA as a template for creating a RNA molecule that
conveys the information encoded in DNA.
In this sense, every gene leads to a unique RNA molecule.
Translation is the molecular biology process by which an RNA molecule's information is
transferred to a protein.
As in transcription, the nucleic acid (DNA in transcription and RNA in translation) serves
as a template for synthesis of the product (RNA in transcription and protein in translation).
Thus, RNA molecules from distinct genes lead to different proteins.

@@

We continue by stating what it means for one trait to mediate a relationship between a
DNA variant and another trait. To clarify our discussion, we refer to an example from
\citet{chick2016defining} (Figure). \citet{chick2016defining}, in
studying livers of 192 Diversity Outbred mice, found evidence that \emph{Dhtkd1}
transcript levels associated with a Chromosome 2 marker near the \emph{Dhtkd1} gene.
They also found that
the same marker affected DHTKD1 protein concentrations. (Note that DHTKD1 protein is the
product of translation of \emph{Dhtkd1} transcripts.)
As anticipated, mediation analysis, in which the DHTKD1 protein concentrations are
regressed on founder allele dosages (at the Chromosome 2 marker) demonstrated that
\emph{Dhtkd1} transcript levels act as a mediator between DHTKD1 protein concentrations
and founder allele dosages. In fact, the extent of the reduction in association
strength indicates that the primary pathway by which the genetic marker affects DHTKD1
protein concentrations is through \emph{Dhtkd1} transcript levels.

@@

To determine whether a local transcript level mediates the relationship between
a nonlocal transcript level and a DNA variant, we perform a series of regression analyses,
which we detail below.
In brief, we regress the nonlocal transcript levels on founder allele dosages at the DNA
variant, with and without conditioning on the candidate mediator (the local transcript levels).
If the LOD score measuring association between nonlocal trait and QTL genotype diminishes
sufficiently upon conditioning on a candidate,
then we declare the candidate a mediator.

@@

@@

%\begin{equ}[!ht]
\begin{equation}
Y = XB + WC + M\beta + E
\label{model4}
\end{equation}
%\caption{Linear model with founder allele dosages, covariates, and candidate mediator.}
%\end{equ}

We now consider the procedures needed for a mediation analysis in systems genetics.
In the four linear regression models, $X$ is an $n$ by $8$ matrix of
founder allele dosages at a single marker, $B$ is an $8$ by $1$ matrix of founder allele
effects, $E$ is an $n$ by $1$ matrix of random errors, $b$
is a number, $1$ is
an $n$ by $1$ matrix with all entries set to $1$, $Y$ is
an $n$ by $1$ matrix of
phenotype values (for a single trait), and $M$ is an $n$
by $1$ matrix of values for a putative mediator.
$C$ is a $c$ by 1 matrix of covariate effects, and $W$ is
a $n$ by $c$ matrix of covariates.
We denote the coefficient of the mediator by $\beta$.

@@

  \begin{enumerate}
\item No unmeasured confounding of the DNA variant-nonlocal transcript levels relationship
\item No unmeasured confounding of the local transcript levels-nonlocal transcript levels relationship
\item No unmeasured confounding of the DNA variant-local transcript levels relationship
\item No local transcript levels-nonlocal transcript levels confounder that is affected by the DNA variant
\end{enumerate}

Our first statistical challenge is due to confounding. To claim that the LOD
difference statistic reflects a causal relationship, four assumptions about
confounding are needed \citep{vanderweele2015explanation},
yet
it is often difficult or impossible to recognize unmeasured confounders. Unmeasured
confounders, as I use the term here, are unmeasured random variables that confound a causal
relationship between two other random variables.
For example, Figure shows a directed graph. Arrows indicate causal
directions. For example, ``QTL'' causes ``M''. ``M'' designates the mediator (local transcript). ``Y'' denotes the nonlocal transcript
level in this setting, while ``C1'', ``C2'', and ``C3'' are covariates that confound,
respectively, the DNA variant - nonlocal transcript level relationship, the local transcript
level - nonlocal transcript level relationship, and the DNA variant - local transcript level
relationship. Failure to measure C1, C2, or C3 leads to violations of Assumption 1,
Assumption 2, or Assumption 3, respectively.

In studies of Diversity Outbred mice, relatedness is a possible confounder,
yet the linear models (Equations~\ref{model1},~\ref{model2},~\ref{model3},~\ref{model4}) fail to account for the complex relatedness patterns
among Diversity Outbred mice.
In fact, the only covariates in our models are for wave membership and sex.
The 400 mice were shipped from the Jackson Laboratory in four waves of 100 mice.
Because each wave was processed at a different time, I include wave membership
as a covariate with the goal of accounting for wave-related effects.
Other sources of confounding, such as batch effects in the phenotyping, may be
unmeasured.

@@

We designed a study to examine the possible roles for mediation analysis and pleiotropy testing.
Because \citet{keller2018genetic} reported that some nonlocal traits that map to
the Chromosome 2 hotspot did not demonstrate evidence of mediation by \emph{Hnf4a} expression
levels, we elected to study a collection of local gene expression traits,
rather than \emph{Hnf4a} alone.
This strategy enabled us to ask whether one of twelve other local traits mediates those
nonlocal hotspot traits that are not mediated by \emph{Hnf4a}.
Our set of local gene expression traits includes \emph{Hnf4a} and 12
other local genes (Table \ref{tab:annot}).
Our 13 local genes are the only genes that met three criteria.
The 147 nonlocal traits that we studied all had LOD peak heights above 7.18 and QTL positions within 2 Mb of the center of the hotspot (at 165.5 Mb).

\begin{enumerate}
    \item QTL peak with LOD $>$ 40
    \item QTL peak position within 2 Mb of hotspot center (165.5 Mb)
    \item Gene midpoint within 2 Mb of hotspot center (165.5 Mb)
\end{enumerate}

@@

Our bivariate QTL analyses involved the same 13 local expression traits and 147 nonlocal expression traits.
We described above the criteria for choosing these expression traits.

@@

We also examined the 1,911 pairs by considering one nonlocal gene at a time.
For each of the 147 nonlocal genes, I plotted LOD difference proportion
against pleiotropy test statistic values.
I present below (Figure ) examples that illustrate some of the
observed patterns between pleiotropy test statistics and LOD difference proportion values.

@@

%\subsection{Scatter plot for all 1911 pairs}
Figure is a scatterplot of the results from all pleiotropy tests and mediation analyses.
Each point contains one local expression trait and one nonlocal expression trait.
I see that points with high values of LOD difference proportion tend to have small
values of pleiotropy test statistic, and those points with high values of the
pleiotropy test statistic tend to have small values of LOD difference proportion.

Some points demonstrate large values of pleiotropy test statistic (\emph{eg.}, $>$ 10),
yet still have sizeable LOD difference proportion statistics (Figure).
One possible explanation for such points is that the univariate LOD (LOD$_1$) value is small,
so that even a modest LOD difference gives rise to a sizeable LOD difference proportion value.
A second possible explanation is that the phenotype variances for both phenotypes in the pair
are large enough to skew the null distribution of the test statistic towards larger values.
To distinguish between these two, I would
examine the LOD difference statistics and obtain bootstrap $p$-values.

In Figure, I colored red the points that involve \emph{Hnf4a};
points with other local genes are blue.
The most striking feature of the coloring is that many red points have small
values of the pleiotropy test statistics and very high values of LOD difference
proportion. This is what I expect when \emph{Hnf4a} mediates the relationship
between a nonlocal trait and a QTL; namely, \emph{Hnf4a} and a nonlocal trait
share a pleiotropic QTL and conditioning on \emph{Hnf4a} diminishes the LOD
score for the association between nonlocal trait and QTL.

@@

To more thoroughly examine the pleiotropy test and mediation analysis
relationships across the 13 local genes,
I created 13 plots of LOD difference proportion against pleiotropy test
statistic (Figures).
They reveal common patterns.
First, I see no points in the upper right quadrant of each plot (Figure).
This tells us that those nonlocal genes with high values of pleiotropy test statistic
(when paired with the specified local gene) have low values of LOD difference proportion.
Similarly, those nonlocal genes with high values of LOD difference proportion tend
to have small values of the pleiotropy test statistic.
Finally, some trait pairs demonstrate low values of both
the LOD difference proportion and pleiotropy test statistic.
This observation suggests that, for a given local expression trait, the nonlocal trait
is not mediated by the local expression trait yet it shares a pleiotropic locus.
Low power to resolve multiple nearby QTL may give rise to such findings.

In comparing the \emph{Hnf4a} plot (Figure ) with the other
12 plots (Figure ),
I see that none of the 12 plots in Figure  closely resembles Figure .
\emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22} are closest in appearance to the plot of \emph{Hnf4a}.
However, each of \emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22}
has very few points with LOD difference proportion above 0.5,
while \emph{Hnf4a} has many points with LOD difference proportion above 0.5.

@@

Now that I've examined our results from the perspective of every one of 13 local gene expression traits,
I turn attention to the nonlocal gene expression trait perspective.
I present four nonlocal gene expression trait plots in Figure. I chose
these four from the 147 to demonstrate two of the observed patterns. In a single scatterplot,
I've plotted 13 points, one for each local gene expression trait, paired with a single
nonlocal gene expression trait. Local expression trait \emph{Hnf4a} is colored in red,
while the other local genes are colored in blue. The top row of Figure
features scatterplots for nonlocal expression traits \emph{Abcb4} and \emph{Smim5}. I see
that local expression trait \emph{Hnf4a} has, by far, the greatest value of LOD difference
proportion. In the second row of Figure, I see a different pattern in
the plots for nonlocal expression traits \emph{Tmprss4} and \emph{Gm3095}. Both demonstrate,
for \emph{Hnf4a}, small values for the pleiotropy test statistic and small values of LOD
difference proportion. This tells me that \emph{Hnf4a} shares a pleiotropic locus with both
\emph{Tmprss4} and \emph{Gm3095}, yet none of the 13 local gene expression traits mediates
the relationship between these two nonlocal traits and the QTL.

@@

This report was generated on \Sexpr{Sys.time()} using the following computational environment and dependencies:
